\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-23 coqimpl.tex
Language Design Section, second attempt.

In our Coq plugin, we can achieve such exposure of definition by
exposing the whole definition directly into module type. Doing so we also prohibit overriding because
subtraction doesn't have the same definition as $add$. We require the
users to decorate field as `Final' so that the plugin will proceed with
this special treatment.\YZ{This para is confusing: which kind of equality is used, propositional or judgmental? Where does 'Final' show up in the example?}\EDJreply{I rephrase it and remove the part on taking about meta-theory. Now everything is about judgemental so I don't have to emphasize judgemental. I wanted to make the example small but I guess I can add one `Final' statement into our example. }

@@

To resolve this, we provide family sealing mechanism---basically we allow the users to specify the family/module type, and thus we can have a \textit{locally transparent}, \textit{non-overridable} field located in a \textit{overridable} family. Using the above example, we simply make \mintinline{Coq}{(f : nat → nat, p : monotonic f)} together inside an \textit{overridable} family, but "f" to be non-overridable and transparent on its definition, and thus we can prove "p". Once we want to swap the implementation, we just override the whole ``sealed'' family.

@@

\textbf{Special Fields: ``Extensible'' Inductive Type and Defining Recursor.}
We don't actually support extensible inductive type from \ref{langdesign-req2} in Coq (because that require non-trivial research effort on theory side---i.e. {What is the semantic of extensible inductive type?}---and engineering), but we simulate extensible inductive type using overriding---when extending an inductive type $A$ with new constructors $c$, our plugin will generate a new inductive definition $A'$ with this new $c$ on the fly and feed to Coq.  But still, we must be careful on the exposing data in the family type (module type) for the following field.

@@

Let's look at \cref{fig:plugin-example2}, the example of boolean and negation. When we define inductive type "b", the following "Field neg_handler.ff" can only know "b" is a type (of "Set") and there are at least two constructors for it, just like what we declare in ``the abstracted interface'' module type "b_3". We cannot export the eliminator of "b" inside "b_3" because doing so "neg_handler.ff" (and "ff_24") will not be able to be inherited to the context where "b" is extended with a third constructor, since the existence of recursor alone will assert that there are only two constructors in "b", not ``extensible'' at all!

@@

\textbf{Propositional Partial Recursor.}
Apparently, the exposed module "b_3" in \cref{fig:plugin-example2} is not expressive enough to really fulfill \ref{langdesign-req2}. For example, we cannot prove the constructors are injective, i.e. \mintinline{Coq}{self_B.tt ≠ self_B.ff}. In the original setting, to prove injectiveness of constructor of (vanilla) inductive type, we only require a recursor and its computational rules. However, as we pointed out earlier, the recursor of an inductive type cannot be exposed into the context for the sake of extensibility. Thus we need to alter the formulation of the recursor i.e., we need to have a better characterization of ``extensible inductive type'' exported in the module type as interface.

@@

In fact, the above "reflect" function implies that propositional partial recursor is good enough to some extent, at least for non-indexed inductive type, because we can ``embed'' vanilla inductive type into types supporting partial recursor. Because "reflect" is actually a left inverse of an injection "bool → self_B.b". A bit more formally, and restrict our focus to non-indexed inductve types,

@@

This style of ``Global Reasoning'' is that we can only prove the corresponding theorem when we know the ``full pictures'', especially the inductive type. We provide "Closing Fact" as a command to carry out this global reasoning---we can attach "Closing Fact" with either (proof) terms or Coq's proof script, then only when compilation to module happens, the term/proof script will put into type-check to see if a "Closing Fact" statement can be proved. Note that, "Closing Fact" cannot immeidately check the correctness of the term/proof script because this verification only happens when we finished defining a family, and thus a bit harder to handle.

@@

\textbf{Hack: ``Meta-data''.} Due to our implementation of the family, the family data is invisible for Coq internal and thus cause some inconvenience.  For example, currently it is not possible to define a vanilla inductive type dependent on our ``extensible'' inductive type due to the invisiblity, unless we compile the whole family. Similarly, customized tactic expression requested by \ref{langdesign-req5} cannot use lemmas or theorems in the fields of the current defining family as well.

To remedy this, we facilitate a "MetaData" command. This command will allow to bundle any original Coq primitive (in the current family context) into the module type, and thus visible to the following defining field. Doing so, we can define customized tactic as in \ref{langdesign-req5} that uses lemmas proved in the family.

>>>>

In our Coq plugin, we can achieve such exposure of definition by
exposing the whole definition directly into the compiled module type. Doing so we also prohibit overriding because
subtraction doesn't have the same definition as $add$. We require the
users to decorate field as `Final' so that the plugin will proceed with
this special treatment.\YZ{This para is confusing: which kind of equality is used, propositional or judgmental? Where does 'Final' show up in the example?}\EDJreply{I rephrase it and remove the part on taking about meta-theory. Now everything is about judgemental so I don't have to emphasize judgemental. I wanted to make the example small but I guess I can add one `Final' statement into our example. }

@@

To resolve this, we provide family sealing mechanism---basically we allow the users to specify the family/compiled module type, and thus we can have a \textit{locally transparent}, \textit{non-overridable} field located in a \textit{overridable} family. Using the above example, we simply make \mintinline{Coq}{(f : nat → nat, p : monotonic f)} together inside an \textit{overridable} family, but "f" to be non-overridable and transparent on its definition, and thus we can prove "p". Once we want to swap the implementation, we just override the whole ``sealed'' family.

@@

\textbf{Special Fields: ``Extensible'' Inductive Type and Defining Recursor.}
We don't actually support extensible inductive type from \ref{langdesign-req2} in Coq (because that require non-trivial research effort on theory side---i.e. {What is the semantic of extensible inductive type?}---and engineering), but we simulate extensible inductive type using overriding---when extending an inductive type $A$ with new constructors $c$, our plugin will generate a new inductive definition $A'$ with this new $c$ on the fly and feed to Coq.  But still, we must be careful on the exposing data in the family type (the compiled module type) for the following field.

@@

Let's look at \cref{fig:plugin-example2}, the example of boolean and negation. When we define inductive type "b", the following "Field neg_handler.ff" can only know "b" is a type (of "Set") and there are at least two constructors for it, just like what we declare in ``the abstracted interface'' the compiled module type "b_3". We cannot export the eliminator of "b" inside "b_3" because doing so "neg_handler.ff" (and "ff_24") will not be able to be inherited to the context where "b" is extended with a third constructor, since the existence of recursor alone will assert that there are only two constructors in "b", not ``extensible'' at all!

@@

\textbf{Propositional Partial Recursor.}
Apparently, the exposed module "b_3" in \cref{fig:plugin-example2} is not expressive enough to really fulfill \ref{langdesign-req2}. For example, we cannot prove the constructors are injective, i.e. \mintinline{Coq}{self_B.tt ≠ self_B.ff}. In the original setting, to prove injectiveness of constructor of (vanilla) inductive type, we only require a recursor and its computational rules. However, as we pointed out earlier, the recursor of an inductive type cannot be exposed into the context for the sake of extensibility. Thus we need to alter the formulation of the recursor i.e., we need to have a better characterization of ``extensible inductive type'' exported in the compiled module type as a part of the interface for the following fields to use.

@@

In fact, the above "reflect" function implies that propositional partial recursor is good enough to some extent, at least for non-indexed inductive type, because we can ``embed'' vanilla inductive type into types supporting partial recursor. Because "reflect" is actually a left inverse of an injection "bool → self_B.b". A bit more formally, and restrict our focus to non-indexed inductive types,

@@

This style of ``Global Reasoning'' is that we can only prove the corresponding theorem when we know the ``full pictures'', especially the inductive type. We provide "Closing Fact" as a command to carry out this global reasoning---we can attach "Closing Fact" with either (proof) terms or Coq's proof script, then only when compilation to module happens, the term/proof script will put into type-check to see if a "Closing Fact" statement can be proved. Note that, "Closing Fact" cannot immediately check the correctness of the term/proof script because this verification only happens when we finished defining a family, and thus a bit harder to handle.

@@

\textbf{Hack: ``Meta-data''.} Due to our implementation of the family, the family data is invisible for Coq internal and thus cause some inconvenience.  For example, currently it is not possible to define a vanilla inductive type dependent on our ``extensible'' inductive type due to the invisibility, unless we compile the whole family. Similarly, customized tactic expression requested by \ref{langdesign-req5} cannot use lemmas or theorems in the fields of the current defining family as well.

To remedy this, we facilitate a "MetaData" command. This command will allow to bundle any original Coq primitive (in the current family context) into the compiled module type, and thus visible to the following defining field. Doing so, we can define customized tactic as in \ref{langdesign-req5} that uses lemmas proved in the family.

### 2022-06-23 lang-design.tex
Save
Language Design Section, second attempt.

@@

>>>>

\subsection{Design Space}

@@

\subsection{Language Design}
Inspired by our above requirements and Family Polymorphism, we propose the following extended syntax for Coq Vernacular Commands. 
% basically write down the specification of surface Coq commandline
%   and then explain  how the language design makes the programmer’s life easier
\begin{align*}
  V := \quad &"Family"~\langle ident\rangle~["extends"~\langle ident\rangle] 
  \quad |\quad "EndFamily" \\
  & |\quad "FRecursor"~\langle ident\rangle~"about"~\langle ident\rangle~"motive"~\langle term\rangle~"using"~\langle ident\rangle \\
  & |\quad "Extend"~"Family"~\langle ident\rangle \\ 
  & |\quad ["Extend"]~"FInductive"~\langle ident\rangle \\ 
  & |\quad ["Final" | "Override"]~"Field"~\langle ident\rangle~":"~\langle term \rangle~":="~\langle term \rangle \\
  & | \quad "Inherit"~\langle ident\rangle \quad | \quad "Inherit Until"~\langle ident\rangle
\end{align*}

We use "Family" and "EndFamily" to scope. Each field in one family is declared and defined by "Field". We can define nested family. Family will also be used to encode recursive handlers. Each case handling of one inductive type will be aggregated into one family, then "FRecursor" will be used to construct a recursor using these handlers. For example, \mintinline{Coq}{FRecursor recur about T motive f} "using" "handlers" will generate a new field "recur" that is doing induction (using elimination principle of "T") using motive "f", and the handlers are located in the family "handlers".

We can inherit a top level family "parent" using \mintinline{Coq}{Family ... extends parent} as a new top level family. As expected, "Override" will override/a field from the parent. "Inherit" can inherit nested family, inductive type and fields. The fields inside a family are order-sensitive because we have dependent type here. So "Inherit Until" is used if the user want to inherit a range of fields. If the user wants to extend an inductive type or a nested family from the parent, then "Extend FInductive" or "Extend Family" is used. Thus nested family can only be inherit or extended together with its outer family (and thus top level family), thus our support of family polymorphism is incomplete.

This incomplete support can already be helpful for extending functions. Take \cref{fig:plugin-example2} as example, if we want to extend boolean inductive type and its negation, we only need to specify the new constructor for boolean and how negation deal with the new constructor.
